Practical Open-Loop Optimistic Planning

### 2019-04-13 paper/main.tex
Updates from Overleaf

\emph{Monte-Carlo Tree Search} (\texttt{MCTS}) algorithms were historically motivated by the application of computer Go, and made a first appearance in the CrazyStone software \citet{Coulom2006}. They were later reformulated in the setting of Multi-Armed Bandits by \citet{Kocsis2006} with their \emph{Upper Confidence bounds applied to Trees} (\texttt{UCT}) algorithm. Despite its popularity, \texttt{UCT} has been shown to suffer from several limitations: its sample complexity can be at least doubly-exponential for some problems (e.g. when a narrow optimal path is hidden in a suboptimal branch), which is much worse than uniform planning \citep{Coquelin2007}. The \texttt{Sparse Sampling} algorithm of \citet{Kearns2002} achieves better worst-case performance, but it is still non-polynomial and doesn't adapt to the structure of the MDP. In stark contrast, the \emph{Optimistic Planning for Deterministic systems} (\texttt{OPD}) algorithm considered by \citet{Hren2008} in the case of deterministic transitions and rewards exploits the structure of the cumulative discounted reward to achieve a problem-dependent polynomial bound on sample complexity. A similar line of work in a deterministic setting is that of \texttt{SOOP} and \texttt{OPC} by \cite{Busoniu2013,Busoniu2018} thoush they focus on continuous action spaces. \texttt{OPD} was later extended to stochastic systems with the \emph{Open-Loop Optimistic Planning} (\OLOP) algorithm introduced by \citet{Bubeck2010} in the open-loop setting: we only consider sequences of actions independently of the states that they lead to. This restriction in the space of policies causes a loss of optimality, but greatly simplifies the planning problem in the cases where the state space is large or infinite. More recent work such as \texttt{St0p} \citep{Szorenyi2014} and \texttt{TrailBlazer} \citep{Grill2016} focus on the probably approximately correct (PAC) framework: rather than simply recommending an action to maximise the expected rewards, they return an $\epsilon$-approximation of the value at the root that holds with high probability. This highly demanding framework puts a severe strain on these algorithms that were developed for theoretical analysis only and cannot be applied to real problems.

@@

Under these circumstances, the Lemma \ref{lemma:seq_values} has a drastic effect on the search behaviour. Indeed, as long as a subtree under the root verifies $U^{\mu}_a(m) > 1$ for every sequence $a$, then all these sequences share the same B-value $B_a(m) = U_{a_{1:1}(m)}$. This means that \OLOP cannot differentiate them and exploit information from their shared history as intended, and behaves as uniform sampling instead.
Once the early depths have been explored sufficiently, \OLOP resumes its intended behaviour, but the problem is only shifted to deeper unexplored subtrees.

@@

\begin{figure}[pht]
    \centering
    \includesvg[width=0.64\textwidth]{img/OLOP}
    \includesvg[width=0.64\textwidth]{img/KL-OLOP}
    \includesvg[width=0.64\textwidth]{img/ODP}
    \caption{The look-ahead trees expanded by different planning algorithms for a budget $n=10^3$, shown down to depth 6. The width of edges is proportional to the nodes visit count $T_a(M)$.}
    \label{fig:trees}
\end{figure}

>>>>

\emph{Monte-Carlo Tree Search} (\texttt{MCTS}) algorithms were historically motivated by the application of computer Go, and made a first appearance in the CrazyStone software \citet{Coulom2006}. They were later reformulated in the setting of Multi-Armed Bandits by \citet{Kocsis2006} with their \emph{Upper Confidence bounds applied to Trees} (\texttt{UCT}) algorithm. Despite its popularity, \texttt{UCT} has been shown to suffer from several limitations: its sample complexity can be at least doubly-exponential for some problems (e.g. when a narrow optimal path is hidden in a suboptimal branch), which is much worse than uniform planning \citep{Coquelin2007}. The \texttt{Sparse Sampling} algorithm of \citet{Kearns2002} achieves better worst-case performance, but it is still non-polynomial and doesn't adapt to the structure of the MDP. In stark contrast, the \emph{Optimistic Planning for Deterministic systems} (\texttt{OPD}) algorithm considered by \citet{Hren2008} in the case of deterministic transitions and rewards exploits the structure of the cumulative discounted reward to achieve a problem-dependent polynomial bound on sample complexity. A similar line of work in a deterministic setting is that of \texttt{SOOP} and \texttt{OPC} by \cite{Busoniu2013,Busoniu2018} though they focus on continuous action spaces. \texttt{OPD} was later extended to stochastic systems with the \emph{Open-Loop Optimistic Planning} (\OLOP) algorithm introduced by \citet{Bubeck2010} in the open-loop setting: we only consider sequences of actions independently of the states that they lead to. This restriction in the space of policies causes a loss of optimality, but greatly simplifies the planning problem in the cases where the state space is large or infinite. More recent work such as \texttt{St0p} \citep{Szorenyi2014} and \texttt{TrailBlazer} \citep{Grill2016} focus on the probably approximately correct (PAC) framework: rather than simply recommending an action to maximise the expected rewards, they return an $\epsilon$-approximation of the value at the root that holds with high probability. This highly demanding framework puts a severe strain on these algorithms that were developed for theoretical analysis only and cannot be applied to real problems.

@@

Under these circumstances, the Lemma \ref{lemma:seq_values} has a drastic effect on the search behaviour. Indeed, as long as a subtree under the root verifies $U^{\mu}_a(m) > 1$ for every sequence $a$, then all these sequences share the same B-value $B_a(m) = U_{a_{1:1}(m)}$. This means that \OLOP cannot discriminate them and exploit information from their shared history as intended, and behaves as uniform sampling instead.
Once the early depths have been explored sufficiently, \OLOP resumes its intended behaviour, but the problem is only shifted to deeper unexplored subtrees.

@@

\begin{figure}[pht]
    \centering

    \includesvg[width=0.64\textwidth]{img/ODP}
    \includesvg[width=0.64\textwidth]{img/OLOP}
    \includesvg[width=0.64\textwidth]{img/KL-OLOP}
    \caption{The look-ahead trees expanded by different planning algorithms for a budget $n=10^3$, shown down to depth 6. The width of edges is proportional to the nodes visit count $T_a(M)$.}
    \label{fig:trees}
\end{figure}
